Data analysis report

### 2024-09-20 notebooks/04-response-letter.qmd

---
title: Response to reviewers
subtitle: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
format: docx
author: "-"
---

### 2024-09-20 notebooks/bayesian-modelling.qmd
Starting to write the manuscript for Word
- Highlight extension by mcanouil
- _quarto.yml : Adjusted the author list, different for the manuscript and analyses reports
- Transferred Overleaf references
- Added protocol figures
- Integrated Bayesian functions and code to main, need to transfer the bayesian into the modelling report
- Added the old reports temporarily for transferring

---
title: Bayesian Modelling
subtitle: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
format: html
---

>>>>

---
title: Bayesian Modelling
subtitle: "Are there unconscious visual images in aphantasia? Development of an implicit priming paradigm"
author: 
  - name: Maël Delem
    orcid: 0009-0005-8518-1991
    email: mael.delem@univ-lyon2.fr
    url: https://emc.univ-lyon2.fr/equipes/memoire-emotion-et-attention-mea/delem-mael
    affiliations:
    - id: emc
      name: Study of Cognitive Mechanisms (EMC) Laboratory
      city: Lyon
      country: France
      url: https://emc.univ-lyon2.fr
---

### 2024-09-20 notebooks/data-analysis-report.qmd
Starting to write the manuscript for Word
- Highlight extension by mcanouil
- _quarto.yml : Adjusted the author list, different for the manuscript and analyses reports
- Transferred Overleaf references
- Added protocol figures
- Integrated Bayesian functions and code to main, need to transfer the bayesian into the modelling report
- Added the old reports temporarily for transferring

# The package `librairian` will ease the package management with the "shelf" 
# function, which automatically: 
# 1) checks if a package is installed 
# 2) installs it if need be
# 3) loads the package like the "library()" function would.
if (!require(librarian)) install.packages(librarian)
library(librarian)

@@

@@

  # ─── essential package collections ───
  easystats,      # data analysis framework
  tidymodels,     # modelling framework
  tidyverse       # modern R ecosystem
)

# ─── Fixing a seed for reproducibility ───
set.seed(14051998)
```

@@

```{r}
#| label: fitting-glm-questionnaires
#| code-summary: "Fitting GLMs on questionnaire scores with `tidymodels`"

@@

# Run calculations for a range of effect sizes
(simulations <-
  foreach (
    effectsize = c(10, 15, 20, 25, 30, 35, 40), 
    .packages = c("tidyverse", "simr")
    ) %dopar% {
      # Calculate regression coefficients corresponding to the effect size and 
      # update the model to simulate from accordingly
      fixef(model) <- extract_coefficients(effectsize)

@@

```{r}
# ═══ Accuracy outliers analysis ═══════════════════════════════════════════════

df_0e <- df_explicit
df_0i <- df_implicit
```

@@

# ─── Preparing tables for correlation ───
congruence_effects_emm <-
  list(
      df_3e = df_3e,
      df_3i = df_3i
      ) |> 
  imap(
    ~.x |> 
      group_by(subjectid, congruence, color) |> 
      summarise(mean_rt = mean(rt)) |> 
      group_by(subjectid, congruence) |> 
      summarise(mean = mean(mean_rt)) |> 
      pivot_wider(
        names_from = congruence,
        values_from = mean
      ) |> 
      mutate(congruence_effect = uncongruent - congruent, .keep = "unused") |> 
      left_join(df_questionnaires[,c(1,5:8)], by = "subjectid") |> 
      ungroup()
    )
```

@@

>>>>

# The package `librairian` will ease the package management with the "shelf" 
# function, which automatically: 
# 1) checks if a package is installed 
# 2) installs it if need be
# 3) loads the package like the "library()" function would.
if (!require("librarian")) install.packages("librarian")
library("librarian")

@@

  # ─── bayesian ───────────────────────
  brms,           # bayesian models
  bayesplot,      # bayesian plots

@@

  # ─── essential package collections ───
  tidymodels,     # modelling framework
  easystats,      # data analysis framework
  tidyverse       # modern R ecosystem
)

# ─── Fixing a seed for reproducibility ───
set.seed(14051998)

# ─── Bayesplot theme ───
color_scheme_set("teal")

# ─── Global theme ───
theme_set(theme_modern(base_size = 14))
```

@@

```{r}
#| label: fitting-glm-questionnaires
#| code-summary: "Fitting GLMs on questionnaire scores with `tidymodels`"
#| eval: false

@@

# Run calculations for a range of effect sizes
(simulations <-
  foreach(
    effectsize = c(10, 15, 20, 25, 30, 35, 40), 
    .packages = c("tidyverse", "simr")
    ) %dopar% {
      # Calculate regression coefficients corresponding to the effect size and 
      # update the model to simulate from accordingly
      fixef(model) <- extract_coefficients(effectsize)

@@

```{r}
#| label: preparing-rt-data

# ═══ Accuracy outliers analysis ═══════════════════════════════════════════════

df_0e <- 
  df_explicit |> 
  mutate(across(c(aphantasia, color, congruence), as.factor))

df_0i <- df_implicit |> 
  mutate(across(c(aphantasia, color, congruence), as.factor))
```

@@

# ─── Preparing tables for correlation ───
congruence_effects_emm <-
  list(
      df_3e = df_3e,
      df_3i = df_3i
      ) |> 
  imap(
    ~.x |> 
      group_by(subjectid, congruence, color) |> 
      reframe(mean_rt = mean(rt)) |> 
      group_by(subjectid, congruence) |> 
      reframe(mean = mean(mean_rt)) |> 
      pivot_wider(
        names_from = congruence,
        values_from = mean
      ) |> 
      mutate(congruence_effect = uncongruent - congruent, .keep = "unused") |> 
      left_join(df_questionnaires[,c(1,5:8)], by = "subjectid") |> 
      ungroup()
    )
```

@@

### Setting contrasts

```{r}
#| label: setting-contrasts

# ─── Setting contrasts ────────────────────────────────────────────────────────
contrasts(df_3e$aphantasia) <- contr.sum(2)
contrasts(df_3e$congruence) <- contr.sum(2)
contrasts(df_3e$color) <- contr.sum(2)
```

### Bayesian news

```{r}
#| label: parallel-setup

# finding the available cores for parallel processing
n_cores <- parallel::detectCores() - 1
```

#### Formula and priors

We used the `brms` R package to fit Bayesian models on the RT data. The package allows for the specification of priors for each parameter of the model. The prior for the intercept was set to a normal distribution with a mean of -0.3 (~0.74s in log-units) and a wide SD of 0. As we expected effect sizes to be inferior to 50ms, we chose weakly informative priors on the coefficients accordingly with normal distributions of mean -2 and wide SDs of 0.

```{r}
#| label: model-brm-formula-priors

brm_formula <- 
  bf(rt ~ aphantasia * congruence * color + (congruence + color | subjectid))
# brm_formula_e <- 
#   bf(rt ~ aphantasia * congruence * color + (1 | subjectid),
#      ndt   ~ (1 | subjectid))

# get_prior(brm_formula_e, data = df_3e, family = shifted_lognormal())

# these priors are weakly informative and were flexible enough to allow for 
# convergence on the (1 | subject) model with ndt per subject, or the
# (congruence + color | subject) model without ndt per subject
brm_priors <- c(
  prior(normal(0, 0.5),  class = b),
  prior(normal(-0.8, 0.5), class = Intercept),
  # prior(normal(-0.7, 0.5), class = Intercept, dpar = ndt),
  prior(normal(1, 0.5), class = ndt),
  prior(normal(0.3, 0.5), class = sd),
  # prior(normal(0.3, 0.5), class = sd, dpar = ndt),
  prior(normal(2, 0.5),   class = sigma)
  )
```

#### Explicit task

##### Prior predictive check

```{r}
#| label: model-brm-explicit-prior-fitting
#| eval: false

n_chains  <- 1

brm_model_priors_e <-
  brm(
    formula = brm_formula, 
    data = df_3e, 
    family = shifted_lognormal(),
    prior = brm_priors,
    chains = n_chains,
    cores  = n_chains,
    warmup = 1000,
    iter = 2000,
    seed = 140598,
    silent = 0,
    refresh = 1,
    file = "analyses-results/brm-explicit-model-priors.rds",
    file_refit = "on_change",
    sample_prior = "only"
    )

pp_check(
  brm_model_priors_e, 
  ndraws = 50, 
  group = "aphantasia",
  type = "dens_overlay_grouped"
  ) +
  scale_x_continuous(
    name = "RT (s)",
    expand = c(0, 0),
    limits = c(0, 3),
    breaks = breaks_pretty(8)
  )
```

##### Model fitting and checks

```{r}
#| label: model-brm-explicit-fitting
#| eval: false

# n_chains  <- 1  # for testing
n_chains  <- n_cores
# warmup <- 100
# iter   <- 200
warmup <- 2000
iter   <- 4200

# init_ndt  <- list(Intercept_ndt = -3)
# inits_ndt <- replicate(n_chains, init_ndt, simplify = FALSE)

brm_model_e <-
  brm(
    formula = brm_formula, 
    data = df_3e, 
    family = shifted_lognormal(),
    prior = brm_priors,
    chains = n_chains,
    cores  = n_chains,
    warmup = warmup,
    iter = iter,
    # init = inits_ndt,
    seed = 140598,
    silent = 0,
    refresh = 1,
    control = list(
      adapt_delta = 0.9,
      max_treedepth = 15
      ),
    # all parameters need to be saved for potential estimation (BFs)
    save_pars = save_pars(all = TRUE),
    file = "analyses-results/brm-model-explicit.rds",
    file_refit = "always"
    )

# 18m 54s for 4200 iterations, full model without ndt

conditional_effects(brm_model_e, effects = "aphantasia:congruence")

pp_check(
  brm_model_e, 
  ndraws = 50, 
  group = "aphantasia",
  type = "dens_overlay_grouped"
  ) +
  scale_x_continuous(
    name = "RT (s)",
    expand = c(0, 0),
    limits = c(0, 3),
    breaks = breaks_pretty(8)
  )
```

##### Model description

```{r}
#| label: model-brm-explicit-description
#| eval: false

summary(brm_model_e)
describe_posterior(brm_model_e, rope_range = c(-0.01, 0.01))
```

##### Model estimates and contrasts

```{r}
#| label: model-brm-explicit-estimate-means
#| eval: false

brm_model_e_e_emm <- 
  emmeans(
    brm_model_e, 
    ~ congruence | aphantasia,
    epred = TRUE
    ) 
# brm_model_e_e_emm |> contrast() 
brm_model_e_e_emm |> pairs()
```

#### Implicit task

##### Prior predictive check

```{r}
#| label: model-brm-formula-priors-2

brm_formula <- 
  bf(rt ~ aphantasia * congruence * color + (congruence + color | subjectid))
# brm_formula_e <- 
#   bf(rt ~ aphantasia * congruence * color + (1 | subjectid),
#      ndt   ~ (1 | subjectid))

# get_prior(brm_formula_e, data = df_3e, family = shifted_lognormal())

# these priors are weakly informative and were flexible enough to allow for 
# convergence on the (1 | subject) model with ndt per subject, or the
# (congruence + color | subject) model without ndt per subject
brm_priors <- c(
  prior(normal(0, 0.5),  class = b),
  prior(normal(-1, 0.5), class = Intercept),
  # prior(normal(-0.7, 0.5), class = Intercept, dpar = ndt),
  prior(normal(0.5, 0.5), class = ndt),
  prior(normal(0.3, 0.5), class = sd),
  # prior(normal(0.3, 0.5), class = sd, dpar = ndt),
  prior(normal(1.2, 0.5),   class = sigma)
  )
```

```{r}
#| label: model-brm-implicit-prior-fitting
#| eval: false

n_chains  <- 1

brm_model_priors_i <-
  brm(
    formula = brm_formula, 
    data = df_3i, 
    family = shifted_lognormal(),
    prior = brm_priors,
    chains = n_chains,
    cores  = n_chains,
    warmup = 1000,
    iter = 2000,
    seed = 140598,
    silent = 0,
    refresh = 1,
    file = "analyses-results/brm-implicit-model-priors.rds",
    file_refit = "on_change",
    sample_prior = "only"
    )

pp_check(
  brm_model_priors_i, 
  ndraws = 50, 
  group = "aphantasia",
  type = "dens_overlay_grouped"
  ) +
  scale_x_continuous(
    name = "RT (s)",
    expand = c(0, 0),
    limits = c(0, 3),
    breaks = breaks_pretty(8)
  )
```

##### Model fitting and checks

```{r}
#| label: model-brm-implicit-fitting
#| eval: false

# n_chains  <- 1  # for testing
n_chains  <- n_cores
# warmup <- 100
# iter   <- 200
warmup <- 2000
iter   <- 4200

# init_ndt  <- list(Intercept_ndt = -3)
# inits_ndt <- replicate(n_chains, init_ndt, simplify = FALSE)

brm_model_i <-
  brm(
    formula = brm_formula, 
    data = df_3i, 
    family = shifted_lognormal(),
    prior = brm_priors,
    chains = n_chains,
    cores  = n_chains,
    warmup = warmup,
    iter = iter,
    # init = inits_ndt,
    seed = 140598,
    silent = 0,
    refresh = 1,
    control = list(
      adapt_delta = 0.9,
      max_treedepth = 15
      ),
    # all parameters need to be saved for potential estimation (BFs)
    save_pars = save_pars(all = TRUE),
    sample_prior = "yes",
    file = "analyses-results/brm-model-implicit.rds",
    file_refit = "always"
    )

# 18m 54s for 4200 iterations, full model without ndt

conditional_effects(brm_model_i, effects = "aphantasia:congruence")

pp_check(
  brm_model_i, 
  ndraws = 50, 
  group = "aphantasia",
  type = "dens_overlay_grouped"
  ) +
  scale_x_continuous(
    name = "RT (s)",
    expand = c(0, 0),
    limits = c(0, 3),
    breaks = breaks_pretty(8)
  )
```

##### Model description

```{r}
#| label: model-brm-implicit-description
#| eval: false

summary(brm_model_i)
describe_posterior(brm_model_i, rope_range = c(-0.01, 0.01))
```

##### Model estimates and contrasts

```{r}
#| label: model-brm-implicit-estimate-means
#| eval: false

brm_model_i_emm <- 
  emmeans(
    brm_model_i, 
    ~ congruence | aphantasia,
    epred = TRUE
    ) 
brm_model_i_emm |> pairs()
```

### What was already there
